% !TeX spellcheck = en_US
\documentclass[a4paper,10pt]{article}

\usepackage{xcolor, amssymb, stmaryrd}


\newcommand{\red}[1]{\textcolor{red}{#1}}

\usepackage{a4wide}


\begin{document}
%\pagenumbering{arabic}
\title{\#267 On Higher Order Program Verification via the Weighted Relational Model of Linear Logic\\ \emph{Detailed Answers}}
\date{}
\maketitle


We thank again the reviewers for their remarks. We address here in some more detail a few technical comments.

\begin{itemize}
	\item Reviewer A:
	\begin{itemize}
		\item \red{Does your semantics make it possible to observe similar phenomena for PHORS? }
		
		
		As we observe in Section 2, our semantics, based on power series with commuting variables, is not well-adapted to capture properties of trees: it essentially collapses different trees as soon as they have, for all $n$, the same number of branches of length $n$. For example, at ll. 274-279 we present a regular and a non context-free language having the same generating function. This simplification is useful in this context, as we are interested in the probability of termination, which only depends on the number of choices and not on their order. 
%		In particular, the system of algebraic equations arising from a $\mathrm{PBHORS}^{<\infty}$ o $\mathrm{PBHORS}^{\infty}$ of any order is always equivalent to the system arising from some order 1 PHORS, that, while possibly having a completely different tree, has the same termination probability. 
		Because of this, it seems difficult to expect that this approach could reproduce the collapse at order 4, which appears to generate from finer (and not just probabilistic-theoretic) properties of trees. In this regard, it could be interesting to study the case in which variables do not commute, that would produce a stricter correspondence between generating functions and languages.
	\end{itemize}

\item Reviewer B:
	\begin{itemize}
	
	\item \red{I wonder how to prove that this fixed point is Conway.}
	
The proof is in [24], property 3.
	
	\item \red{What do you mean by the point-wise partial order?}
	
	We indeed consider the pointwise order over the coefficients of the power series. 
	In Theorem 6.7, that the minimal solution of the system is indeed $a_G(1)$ can be shown using the fact that fixpoints in QRel are defined by iteration: given a system of polynomial equations $(x_i=p(\vec x,z))_{i \in I}$, its lfp $a_I(z)$ is the sup of $a_I^0(z)=0$ and $a_I^{n+1}(z)=p(a_I^n(z),z)$, while
	the lfp y of $(x_i=p(\vec x,z))_{i \in I}$ is the sup of $y_I^0=0, y_I^{n+1}=p_I(y^n,1)$; as evaluation in 1 is a Scott continous semiring homomorphism, we can prove by induction over $n$ that $a_I^{n}(1)=y_I^n$, from which it follows that $a_I(1)=y$.
	
		\item\red{ p.9, Def.6.1: What does the final judgment mean? (The same expression also appears in Definition 7.1.)}\\
	The type judgement $\mathcal{N} \mid x_1 \dots x_n \vdash t: o$ refers to the fact the each nonterminal of a PBHORS must have type of the form $!_{k_1}\tau_1  \multimap \dots \multimap !_{k_n } \tau_n \multimap o $ (this requirement is already present in the non-graded case of general PHORS). We took the same notation from [37], Def.~2.9, for uniformity, but we agree that it is not very clear.
	
	\item \red{- p.9, l.993: "in the affine system of (Ong et al.)" Please cite this properly. Is this meant to refer to [37]? If so, calling it "Ong et al." ignores the author order in the paper and is disrespectful to the authors of [37]}
	
	Of course, this is a typo, in all other places in which we cited [37] we indeed used ``Li et al.''. Thanks for spotting this.
	
	\item\red{p.10, l.1066: What do you mean by $ [\![t_G]\!]^{\mathbb{R}^{+\infty}_{\ge 0}} $? The coefficient semiring in the interpretation had always been assumed to be $ \mathbb{R}^{+\infty}_{\ge 0}\{\{ \{z\} + \Sigma \}\} $ (on line 619), but this seems to contradict that. I think a somewhat more careful argument is needed here.}
	
Yes, that is a typo, thanks for spotting that. The coefficient semiring for the interpretation of $t_G$ for $PBHORS^{<\infty}$ is $\mathbb{R}^{+\infty}_{\ge 0}\ \{\{z\}\}$ (l. 834). Moreover, the statement ``all coefficients are rational'' should be made more precise: 
 $t_G$ is a polynomial with rational coefficients and variables $z, \mathcal{N}$. Indeed, as the equations defining the PHORS are of the form $F = t_L \oplus_p t_R$, with $p \in \mathbb{Q}$, we can see that $t_G$ is of the form $ zt'_G$, where $t'_G$ belongs to the polynomial ring $ \mathbb{Q}^{+\infty}_{\ge 0} \{\llbracket \mathcal{N} \rrbracket \} $.	

	\item\red{ p.10, Example 7.1: I do not think this represents the Dyck language (there is obviously a branch corresponding to (ace)).
}

Our explanation of the example is probably not the clearest, we admit. But what we claim is that the branch language contains all words of the form ``$w )$'', where $w$ is a word from the Dyck language. To be more precise, we should have actually written all words of the form ``$w ) e$'', where $w$ is from the Dyck language. So, for instance, the word ``$ace$'', corresponding to ``$)e$'', is indeed part of our claimed branch language.

\item \red{ p.11, Fig.2: Please explain the $@_{\infty}$ rule in a bit more detail. Since the distinction between finite and infinitary types is quite powerful, it seems to me that the argument in Section 7.2 would still go through even without this disjointness assumption.}

Thanks for this observation. We realized indeed that Theorem 7.4 can hold without the disjointness condition.

\item\red{p.10, Remark 6.3: I believe that the coefficients of $a_G$ themselves can be computed directly and fairly easily}
	
	This is true, but it should noted that the formulas in [6] can be used to make asymptotic estimates that are not possible by a direct computation. In particular, according to Theorem 3 in [6], one could obtain estimates of the form $\mathbb{P}(\mathcal{G}\downarrow_n) \sim n^\alpha \rho^n $ for some $\alpha, \rho \in \mathbb{R}$. While the computation of these constants is not treated in this paper, the approach through fixpoint algebraic systems opens the way to this kind of analysis, which would not be possible via direct computations.
	

\end{itemize}

\end{itemize}


\end{document}


